---
title: "Pseudocercospora epidemiology"
author: "Aaron Weed"
date: "2026-08-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(tidyverse)
library(ggpubr)
library(GGally)
```

## Overview

Cheyenne would like to model:

- Spore counts as a function of environmental factors (temp, rain, RH): 
-- Sporecounts_Master.xlsx: she collected spore count data at two sites over a few years
-- Model system in avocado predicted spore counts by avg weekly temperature and total (weekly) rainfall

- Disease severity by treatment 
-- Pseudocercospora_disease_severity_Aug2026.xlsx: 


## Spore Counts

### Import and Examine the data 

```{r spore_counts, fig.width= 8, fig.height= 8, warning=FALSE, comment= FALSE, include=TRUE}

Sporecounts_Master <- read_excel("data/Sporecounts_Master.xlsx", 
    sheet = "Master sheet", col_types = c("numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "text", "text", "text", 
        "text", "text", "date", "date"))
View(Sporecounts_Master)

# Numeric variables to plot
numeric_vars <- c("nSpores", "Tmax", "Tmin", "Tave", "Rhm", "Rain")

df_long <- Sporecounts_Master %>%
  pivot_longer(
    cols = all_of(numeric_vars),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  filter(!is.na(Value))

ggplot(df_long, aes(x = Value, fill = Season)) +
  geom_histogram(
    bins = 20,
    alpha = 0.5,
    position = "identity"
  ) +
  facet_grid(
    Variable ~ Orchard*Season,
    scales = "free"
  ) +
  theme_classic() +
  labs(
    x = "Value",
    y = "Frequency",
    fill = "Season"
  )

# Scatter plots

ggscatter(Sporecounts_Master[numeric_vars])

# Compare all orchards and seasons

ggscatter(
  Sporecounts_Master,
  x = "Tave",
  y = "nSpores",
  facet.by = c("Orchard", "Season"),
  add = "reg.line",
  conf.int = TRUE,
  cor.coef = TRUE,
  cor.method = "pearson",
  xlab = "Mean temperature (°C)",
  ylab = "Number of spores"
)

# Scatterplot matrix for the weather variables

vars <- c("nSpores","Tmax", "Tmin", "Tave", "Rhm", "Rain")

Sporecounts_Master %>%
  select(all_of(vars), Orchard, Season) %>%
  ggpairs(
    columns = vars,
    aes(color = Orchard, alpha = 0.5),
    upper = list(
      continuous = wrap("cor", method = "spearman", size = 3.5)
    ),
    lower = list(
      continuous = wrap("points", alpha = 0.4, size = 1.2)
    ),
    diag = list(
      continuous = wrap("densityDiag", alpha = 0.4)
    )
)


# Scatterplots for the weather variables vs Spores

df_long2 <- Sporecounts_Master %>%
  pivot_longer(
    cols = c(Tmax, Tmin, Tave, Rhm, Rain),
    names_to = "WeatherVariable",
    values_to = "WeatherValue"
  ) %>%
  mutate(log_nSpores = log1p(nSpores))

ggplot(df_long2, aes(x = WeatherValue, y = log_nSpores)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_grid(
    WeatherVariable ~ Orchard + Season,
    scales = "free_x"
  ) +
  theme_classic() +
  labs(
    x = "Weather variable",
    y = "log(Number of spores + 1)"
  )


Sporecounts_Master %>%
  select(all_of(weather_vars), Orchard, Season) %>%
  ggpairs(
    columns = weather_vars,
    aes(color = Orchard, alpha = 0.5),
    upper = list(
      continuous = wrap(
        "cor",
        method = "pearson",
        size = 4
      )
    ),
    lower = list(
      continuous = wrap(
        "points",
        alpha = 0.4,
        size = 1.5
      )
    ),
    diag = list(
      continuous = wrap(
        "densityDiag",
        alpha = 0.4
      )
    )
  )

```

### Seasonal Weather

The below plots show weekly weather variables by Orchard and season. 

Ho do you calculate your variable "week". It seem like week 10 (and likley others) varies per year.

```{r spore_weather, fig.width= 8, fig.height= 8, warning=FALSE, comment= FALSE, include=TRUE}

weather_vars <- c("Tmax", "Tmin", "Tave", "Rhm", "Rain")

weather_long <- Sporecounts_Master %>%
  pivot_longer(
    cols = all_of(weather_vars),
    names_to = "Variable",
    values_to = "Value"
  )

# Weather variables by start Date
ggplot(
  weather_long,
  aes(
    x = Start_date,
    y = Value,
    color = Season,
    group = Season
  )
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  facet_grid(
    Variable ~ Orchard,
    scales = "free_y"
  ) +
  theme_classic() +
  labs(
    x = "Date",
    y = NULL,
    color = "Season", main= "Weather variables by Date"
  ) +
  theme(
    strip.text = element_text(size = 10),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    legend.position = "bottom"
  )

# Weather by days since season start

df <- Sporecounts_Master %>%
  mutate(
    StartDate = as.Date(Start_date),
    SeasonStart = as.Date(
      paste0(
        substr(Season, 1, 4),
        "-10-01"
      )
    ),
    DaysSinceStart = as.numeric(StartDate - SeasonStart)
  )


weather_long <- df %>%
  pivot_longer(
    cols = all_of(weather_vars),
    names_to = "Variable",
    values_to = "Value"
  )

ggplot(
  weather_long,
  aes(
    x = DaysSinceStart,
    y = Value,
    color = Season,
    group = Season
  )
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  facet_grid(
    Variable ~ Orchard,
    scales = "free_y"
  ) +
  theme_classic() +
  labs(
    x = "Days since start of season",
    y = NULL,
    color = "Season"
  ) +
  theme(
    legend.position = "bottom"
  )
```



