---
title: "Pseudocercospora epidemiology"
author: "Aaron Weed"
date: "2026-08-21"
output:
  html_document:
    theme: cerulean        # Choose a modern Bootstrap theme
    highlight: tango       # Style your code syntax highlighting
    toc: true              # Add a Table of Contents
    toc_float: true        # Make the Table of Contents float as you scroll
    number_sections: true  # Add numbered headings automatically
    code_folding: hide     # Allow readers to show/hide R code blocks
    df_print: paged        # Render data frames as beautiful interactive tables
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(tidyverse)
library(ggpubr)
library(knitr)
library(kableExtra)
library(GGally)

styled_kable <- function(x, caption = NULL, digits = 2) {
  
  x %>%
    kable(
            digits = digits,
      caption = caption
    ) %>%
    kable_styling(
      bootstrap_options = c(
        "striped",
        "hover",
        "condensed"
      ),
      full_width = FALSE
    )
}
```

## Overview

**Questions:**

1. When are spores spread in the orchard?

- Does spore release correlate with certain climatic conditions (temperature, relative humidity, rain)?
- Are spores more prevalent during certain stages of kiwifruit development
- Is there a dominant direction that the spores are coming from?

2. When should chemicals be applied during the season?

- At what developmental stages should chemicals be applied to reduce disease severity?
- Is disease severity higher at certain parts of the plant (i.e. cordon, middle or tip of branch)

**Datasets:**

- Spore counts over time in each orchard with weather variables: Sporecounts_Master.xlsx: Cheyenne collected spore count data at two sites over a few years

- There is a similar system in avocado which predicted spore counts by avg weekly temperature and total (weekly) rainfall (Darvas 1982)

- Disease severity by treatment: Pseudocercospora_disease_severity_Aug2026.xlsx: 

## Spore Counts

### Import and Examine the data 

```{r spore_data, fig.width= 12, fig.height= 8, warning=FALSE, comment= FALSE, include=TRUE, results= 'hide', fig.align='left'}

Sporecounts_Master <- read_excel("C:/Users/aweed/OneDrive - DOI/Documents/Projects/Pseudocercospora/data/Sporecounts_Master.xlsx",sheet = "Master sheet",
    col_types = c("numeric", "numeric", "numeric",
        "numeric", "numeric", "numeric",
        "numeric", "numeric", "text", "text",
        "text", "text", "text", "date", "date"))
```

**The dataset contains:**

`r nrow(Sporecounts_Master)` total observations (rows)

Observations from `r unique(Sporecounts_Master$Season)` seasons

N observations from `r Sporecounts_Master %>% count(Season, Orchard)`

### Summary Statistics

```{r spore_summ, fig.width= 12, fig.height= 8, warning=FALSE, comment= FALSE, include=TRUE, fig.align='left'}

# Generate Summary stats per grouping variable (Orchard, Season)

# Identify numeric variables
numeric_vars <- Sporecounts_Master %>%
  select(where(is.numeric)) %>%
  names()

# Summary statistics by Season, Orchard, and year
summary_stats <- Sporecounts_Master %>%
  pivot_longer(
    cols = all_of(numeric_vars),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  group_by(Orchard, Season, Variable) %>%
  summarise(
    N = n(),
    Missing = sum(is.na(Value)),
    Percent_Missing = 100 * mean(is.na(Value)),
    Mean = mean(Value, na.rm = TRUE),
    SD = sd(Value, na.rm = TRUE),
    Median = median(Value, na.rm = TRUE),
    Min = min(Value, na.rm = TRUE),
    Max = max(Value, na.rm = TRUE),
    .groups = "drop"
  )

styled_kable(
  summary_stats,
  caption = "Summary statistics by season and orchard"
)
```


### Data Checks

```{r spore_counts, fig.width= 12, fig.height= 8, warning=FALSE, comment= FALSE, include=TRUE, fig.align='left'}

# Identify numeric variables
numeric_vars <- Sporecounts_Master %>%
  select(where(is.numeric)) %>%
  names()

# Missing values by variable for each

missing_check <- Sporecounts_Master %>% 
  group_by(Orchard, Season) %>% 
  summarise(
    across(
      numeric_vars,
      ~ sum(is.na(.))
    )
  ) %>%
  pivot_longer(
    cols = all_of(numeric_vars),
    names_to = "Variable",
    values_to = "Missing"
  )  %>%
  mutate(
    Percent_Missing = 100 * Missing / nrow(Sporecounts_Master)
  )

  styled_kable(
  missing_check %>% filter(Missing > 0),
  caption = "Missing values by variable"
)

# Check for impossible or suspicious weather values

Sporecounts_Master %>%
  summarise(
    Tmin_below_zero = sum(Tmin < 0, na.rm = TRUE),
    Tmax_below_zero = sum(Tmax < 0, na.rm = TRUE),
    Tmax_less_Tmin = sum(Tmax < Tmin, na.rm = TRUE),
    Rhm_below_0 = sum(Rhm < 0, na.rm = TRUE),
    Rhm_above_100 = sum(Rhm > 100, na.rm = TRUE),
    Rain_below_0 = sum(Rain < 0, na.rm = TRUE),
    nSpores_below_0 = sum(nSpores < 0, na.rm = TRUE)
  ) %>%
  kable(
    format = "html",
    digits = 2,
    caption = "Check for impossible or suspicious weather values"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
week_check <- Sporecounts_Master %>%
  group_by(Season, Week) %>%
  summarise(
    Start_date = min(Start_date, na.rm = TRUE),
    End_date   = max(End_date, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Season, Week) %>%
  group_by(Season) %>%
  mutate(
    Previous_Week = lag(Week),
    Previous_Start = lag(Start_date),
    Days_from_previous = as.numeric(
      Start_date - Previous_Start
    ),
    Week_difference = Week - Previous_Week
  ) %>%
  ungroup()


#identifies cases where consecutive Week values are not seven days apart.

styled_kable(
  week_check %>%
  filter(
    !is.na(Days_from_previous),
    Days_from_previous != 7
  ),
  caption = "consecutive Week values that are not seven days apart."
)

```

### Exploratory Plots

```{r spore_explore_plots, fig.width= 12, fig.height= 8, warning=FALSE, comment= FALSE, include=TRUE, results= 'hide',fig.align='left'}

# Numeric variables to plot
numeric_vars <- c("nSpores", "Tmax", "Tmin", "Tave", "Rhm", "Rain")

df_long <- Sporecounts_Master %>%
  pivot_longer(
    cols = all_of(numeric_vars),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  filter(!is.na(Value))

ggplot(df_long, aes(x = Value, fill = Season)) +
  geom_histogram(
    bins = 20,
    alpha = 0.5,
    position = "identity"
  ) +
  facet_grid(
    Variable ~ Orchard*Season,
    scales = "free"
  ) +
  theme_classic() +
  labs(
    x = "Value",
    y = "Frequency",
    fill = "Season",
    title = "Histograms of measured variables by Orchard and Season"
  )

# Scatter plots

# Scatter Plot matrix of measured variables

Sporecounts_Master %>%
  select(all_of(numeric_vars), Orchard, Season) %>%
  ggpairs(title ="Scatter Plot matrix of measured variables",
    columns = numeric_vars,
    aes(color = Orchard, alpha = 0.5),
    upper = list(
      continuous = wrap("cor", method = "spearman", size = 3.5)
    ),
    lower = list(
      continuous = wrap("points", alpha = 0.4, size = 1.2)
    ),
    diag = list(
      continuous = wrap("densityDiag", alpha = 0.4)
    )
)


# Scatterplots for the weather variables vs Spores

df_long2 <- Sporecounts_Master %>%
  pivot_longer(
    cols = c(Tmax, Tmin, Tave, Rhm, Rain),
    names_to = "WeatherVariable",
    values_to = "WeatherValue"
  ) %>%
  mutate(log_nSpores = log1p(nSpores))

# Create separate plots
weather_plots <- df_long2 %>%
  split(.$WeatherVariable) %>%
  lapply(function(dat) {
    
    ggplot(dat, aes(x = WeatherValue, y = log_nSpores)) +
      geom_point(alpha = 0.5) +
      geom_smooth(
        method = "lm",
        se = FALSE
      ) +
      facet_grid(
        ~ Orchard + Season,
        scales = "free_x"
      ) +
      theme_classic() +
      labs(
        x = unique(dat$WeatherVariable),
        y = "log(Number of spores + 1)",
        title = paste(
          unique(dat$WeatherVariable),
          "vs Spores (log)"
        )
      )
  })

# View the individual plots
weather_plots$Tmax
weather_plots$Tmin
weather_plots$Tave
weather_plots$Rhm
weather_plots$Rain
```

### Seasonal Weather

The below plots show weekly weather variables by Orchard and season. 

```{r spore_weather, fig.width= 12, fig.height= 8, warning=FALSE, comment= FALSE, include=TRUE, results= 'hide',fig.align='left'}

weather_vars <- c("Tmax", "Tmin", "Tave", "Rhm", "Rain")

Sporecounts_Master %>%
  select(all_of(weather_vars), Orchard, Season) %>%
  ggpairs(title = "Scatter Plot Matrix of Weather variables",
    columns = weather_vars,
    aes(color = Orchard, alpha = 0.5),
    upper = list(
      continuous = wrap(
        "cor",
        method = "pearson",
        size = 4
      )
    ),
    lower = list(
      continuous = wrap(
        "points",
        alpha = 0.4,
        size = 1.5
      )
    ),
    diag = list(
      continuous = wrap(
        "densityDiag",
        alpha = 0.4
      )
    )
  )


weather_long <- Sporecounts_Master %>%
  pivot_longer(
    cols = all_of(weather_vars),
    names_to = "Variable",
    values_to = "Value"
  )

# Weather variables by start Date
ggplot(
  weather_long,
  aes(
    x = Start_date,
    y = Value,
    color = Season,
    group = Season
  )
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  facet_grid(
    Variable ~ Orchard,
    scales = "free_y"
  ) +
  theme_classic() +
  labs(
    x = "Date",
    y = NULL,
    color = "Season", main= "Weather variables by Date"
  ) +
  theme(
    strip.text = element_text(size = 10),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    legend.position = "bottom"
  )

# Weather by days since season start

df <- Sporecounts_Master %>%
  mutate(
    StartDate = as.Date(Start_date),
    SeasonStart = as.Date(
      paste0(
        substr(Season, 1, 4),
        "-10-01"
      )
    ),
    DaysSinceStart = as.numeric(StartDate - SeasonStart)
  )


weather_long <- df %>%
  pivot_longer(
    cols = all_of(weather_vars),
    names_to = "Variable",
    values_to = "Value"
  )

ggplot(
  weather_long,
  aes(
    x = DaysSinceStart,
    y = Value,
    color = Season,
    group = Season
  )
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  facet_grid(
    Variable ~ Orchard,
    scales = "free_y"
  ) +
  theme_classic() +
  labs(
    x = "Days since start of season",
    y = NULL,
    color = "Season"
  ) +
  theme(
    legend.position = "bottom"
  )

```

### Regression models

```{r spre_pred,fig.width= 12, fig.height= 8, warning=FALSE, comment= FALSE, include=TRUE, results= 'hide',fig.align='left'}



```


